{\noindent \Huge \textbf{Acknowledgments}}

There are many people that I would like to thank.  First I would like to thank my family for enabling and encouraging me in all of my life's pursuits.  To my grandma, for putting up with my \qw{experiments} in her kitchen when I was younger.  To my grandpa, for the musical training that gave me the arts to complement my sciences.  To my brother\ldots yes, I'm thanking you.  To my dad, for teaching me about working hard and having patience.  To my mom, for teaching me to do things the right way.  And to my wife, your confidence and patience (especially the patience) mean the world to me.

Next, I would like to thank Dr. Dan Palmer for giving me the opportunity to do swarm research while I was an undergraduate at John Carroll University.  Also, I want to thank Dr. Palmer for all of his advice and guidance, both academically and personally.  In the same vein, I want to thank Dr. Palmer and Chad Hantak for having the inspiration and vision to develop the first incarnation of \SWEEP.

I would also like to thank Orbital Research for providing me with the opportunity to get my Masters.  Throughout this process, there have been many supportive people at Orbital, too many to thank right now.  But suffice it to say, thank you all for bettering me as a writer and a researcher, giving me a sounding board for my ideas, and providing me with coffee and encouragement.

I would like to express my thanks to my thesis advisers, Dr. Michael Branicky, Dr. Dan Palmer, and Dr. Ravi Vaidyanathan, for all their time and effort, especially in the last few weeks of this process.  Their guidance and input truly helped me navigate this endeavor.  And to the rest of my committee, Dr. Randy Beer and Dr. Roger Quinn, thank you for taking time out of your schedules to read my thesis and be on my committee.

